---
title: "TrialNet proinsulin ratio analyses"
author: "Laura Pyle"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)


library(dplyr)
library(lubridate)
library(graphics)
library(ggplot2)
library(Hmisc)
#library(Table1)
library(knitr)
library(nlme)
#library(ppcor)
#library(sjlabelled)

# read in the data
longitudinal_visit_info <- read.csv("S:\\Shared Projects\\Laura\\BDC\\Projects\\Andrea Steck\\Trialnet progressors analysis\\Fourth analysis - demethylated insulin with updated dataset\\Data\\longitudinal 051019.csv")
ParticipantInfo_risk <- read.csv("S:\\Shared Projects\\Laura\\BDC\\Projects\\Andrea Steck\\Trialnet progressors analysis\\Fourth analysis - demethylated insulin with updated dataset\\Data\\Participant Info risk 051019.csv",na.strings = c("NA"," ","","      "))

source("C:\\Users\\pylell\\Documents\\GitHub\\General-code\\temp_table1.r")
source("C:\\Users\\pylell\\Documents\\GitHub\\General-code\\01_functions.r")


# get rid of duplicates
ParticipantInfo_risk <- unique(ParticipantInfo_risk)
# no duplicates

obslong <- nrow(longitudinal_visit_info)
nppt <- nrow(ParticipantInfo_risk)

# merge participant info and longitudinal data
keep_ppt <- data.frame(ParticipantInfo_risk$TrialNet,
                       ParticipantInfo_risk$T1D,
                       ParticipantInfo_risk$Date.of.Diagnosis,
                       ParticipantInfo_risk$DOB, ParticipantInfo_risk$Gender, ParticipantInfo_risk$Race, ParticipantInfo_risk$Ethnicity,
                       ParticipantInfo_risk$DPT1.Risk.Score,ParticipantInfo_risk$HLA, ParticipantInfo_risk$DR3,ParticipantInfo_risk$DR4) 
colnames(keep_ppt) <- c("Random_ID","T1D","Date_of_Diagnosis","DOB","Gender",
                        "Race","Ethnicity","DPT1Risk","HLA","DR3","DR4")
keep_long <- data.frame(longitudinal_visit_info$PID,
                        longitudinal_visit_info$Draw.Date,
                        longitudinal_visit_info$GAD65,
                        longitudinal_visit_info$GAD65H,
                        longitudinal_visit_info$ICA512,
                        longitudinal_visit_info$IA2H,
                        longitudinal_visit_info$mIAA,
                        longitudinal_visit_info$ZnT8,
                        longitudinal_visit_info$ICA,
                        longitudinal_visit_info$ECL_GADA,
                        longitudinal_visit_info$ECL_IA.2A,
                        longitudinal_visit_info$ECL_IAA,
                        longitudinal_visit_info$Proinsulin_pM.L,
                        longitudinal_visit_info$PEP0,
                        longitudinal_visit_info$HbA1c,
                        longitudinal_visit_info$GLU0,
                        longitudinal_visit_info$GLU30,
                        longitudinal_visit_info$GLU60,
                        longitudinal_visit_info$GLU90,
                        longitudinal_visit_info$GLU120,
                        longitudinal_visit_info$Height.cm,
                        longitudinal_visit_info$Weight.kg)
colnames(keep_long) <- c("Random_ID","Draw_Date","GAD65","GAD65H","ICA512","IA_2H","mIAA","ZnT8","ICA","ECL_GADA","ECL_IA2A","ECL_IAA","Proinsulin_pM_L","PEP0","HbA1c","GLU0","GLU30","GLU60","GLU90","GLU120","Height.cm","Weight.kg")

# convert draw date to date
#keep_long$Draw_Date <- mdy_hms(as.character(keep_long$Draw_Date))
#keep_long$Draw_Date <- as.POSIXlt(as.character(keep_long$Draw_Date),format="%m%d%Y %H%M")
keep_long$Draw_Date <- mdy_hm(as.character(keep_long$Draw_Date))
keep_ppt$DOB <- mdy(as.character(keep_ppt$DOB))

# calculate fasting Cpeptide in pM/L
keep_long$PEP0_pM_L <- keep_long$PEP0*331.09

# calculate proinsulin/Cpep ratio
keep_long$Pro_pep <- keep_long$Proinsulin_pM_L/keep_long$PEP0_pM_L*100

# merge  datasets
single_gada_miaa <- merge(keep_ppt,keep_long,by="Random_ID",all.x=F,all.y=F)
N <- length(unique(single_gada_miaa$Random_ID))

# fix gender coding female Female   male   Male 
single_gada_miaa$Gender[single_gada_miaa$Gender=='female'] <- "Female"
single_gada_miaa$Gender[single_gada_miaa$Gender=='male'] <- "Male"
single_gada_miaa$Gender <- droplevels(single_gada_miaa$Gender)

# new variable for DR3/DR4
single_gada_miaa$DR3DR4[single_gada_miaa$DR3=="PRESENT" & single_gada_miaa$DR4=="PRESENT"] <- "PRESENT"
single_gada_miaa$DR3DR4[!(single_gada_miaa$DR3=="PRESENT" & single_gada_miaa$DR4=="PRESENT")] <- "ABSENT"
single_gada_miaa$DR3DR4[is.na(single_gada_miaa$DR3) & is.na(single_gada_miaa$DR4)] <- NA

# correct some values per Andrea
single_gada_miaa$Height.cm[single_gada_miaa$Draw_Date==as.Date("2008-12-30") & single_gada_miaa$Random_ID=="146273"] <- 172.5
single_gada_miaa$Weight.kg[single_gada_miaa$Draw_Date==as.Date("2008-12-30") & single_gada_miaa$Random_ID=="146273"] <- 57
single_gada_miaa$Height.cm[single_gada_miaa$Draw_Date==as.Date("2008-11-26") & single_gada_miaa$Random_ID=="146513"] <- 164.6
single_gada_miaa$Weight.kg[single_gada_miaa$Draw_Date==as.Date("2008-11-26") & single_gada_miaa$Random_ID=="146513"] <- 52.9
single_gada_miaa$Height.cm[single_gada_miaa$Draw_Date==as.Date("2013-10-14") & single_gada_miaa$Random_ID=="148088"] <- 161.3
single_gada_miaa$Weight.kg[single_gada_miaa$Draw_Date==as.Date("2013-10-14") & single_gada_miaa$Random_ID=="148088"] <- 67
single_gada_miaa$Height.cm[single_gada_miaa$Draw_Date==as.Date("2010-08-07") & single_gada_miaa$Random_ID=="149118"] <- 156.9
single_gada_miaa$Weight.kg[single_gada_miaa$Draw_Date==as.Date("2010-08-07") & single_gada_miaa$Random_ID=="149118"] <- 61.8
single_gada_miaa$Height.cm[single_gada_miaa$Draw_Date==as.Date("2009-04-20") & single_gada_miaa$Random_ID=="150283"] <- 184.6
single_gada_miaa$Weight.kg[single_gada_miaa$Draw_Date==as.Date("2009-04-20") & single_gada_miaa$Random_ID=="150283"] <- 125.4
single_gada_miaa$Height.cm[single_gada_miaa$Draw_Date==as.Date("2015-04-06") & single_gada_miaa$Random_ID=="237376"] <- 174.9
single_gada_miaa$Weight.kg[single_gada_miaa$Draw_Date==as.Date("2015-04-06") & single_gada_miaa$Random_ID=="237376"] <- 75.39
  
# calculate BMI
single_gada_miaa$BMI <- single_gada_miaa$Weight.kg / ((single_gada_miaa$Height.cm/100)^2)
#bad <- single_gada_miaa[single_gada_miaa$BMI>50,]
#bad <- bad[!is.na(bad$Random_ID),]
#bad <- subset(bad,select=c("Random_ID","Draw_Date","Height.cm","Weight.kg","BMI"))

# combine race/ethnicity
single_gada_miaa$race_eth <- NA
single_gada_miaa$race_eth[(single_gada_miaa$Race=="White" & single_gada_miaa$Ethnicity=="Non-Hispanic origin")]  <- "NHW"
single_gada_miaa$race_eth[(single_gada_miaa$Race=="White" & single_gada_miaa$Ethnicity=="Non-Hispanic White")] <- "NHW"
single_gada_miaa$race_eth[(single_gada_miaa$Race=="White" & single_gada_miaa$Ethnicity == "Not Hispanic or Latino")] <- "NHW"
single_gada_miaa$race_eth[single_gada_miaa$Ethnicity=="Hispanic" | single_gada_miaa$Ethnicity=="Hispanic or Latino" | single_gada_miaa$Ethnicity=="Hispanic/Latino/Spanish origin"] <- "Hispanic"
single_gada_miaa$race_eth[is.na(single_gada_miaa$race_eth)] <- "Other"
single_gada_miaa$race_eth <- as.factor(single_gada_miaa$race_eth)

# first code each antibody as positive negative
for(i in 1:nrow(single_gada_miaa)){
  ###coding for miaa
  a <-ifelse(is.na(single_gada_miaa$mIAA[i]), 
             NA,
             ifelse(single_gada_miaa$mIAA[i] <= 0.01, 0, 1)
  )
  single_gada_miaa$miaa_cat[i]<- a
  ###coding for first form of gad65
  b <-ifelse(is.na(single_gada_miaa$GAD65[i]), 
             NA,
             ifelse(single_gada_miaa$GAD65[i] <= 0.032, 0, 1)
  )
  single_gada_miaa$gad65_cat[i]<- b
  ###coding for second form of gad65
  c<-ifelse(is.na(single_gada_miaa$GAD65H[i]), 
            NA,
            ifelse(single_gada_miaa$GAD65H[i] <= 20, 0, 1)
  )
  single_gada_miaa$gad65h_cat[i]<- c
  
  ###coding for first form of ica512/ia2
  d<-ifelse(is.na(single_gada_miaa$ICA512[i]), 
            NA,
            ifelse(single_gada_miaa$ICA512[i] <= 0.049, 0, 1)
  )
  single_gada_miaa$ica512_cat[i]<- d
  ###coding for second form of ica512/ia2
  e<-ifelse(is.na(single_gada_miaa$IA_2H[i]), 
            NA,
            ifelse(single_gada_miaa$IA_2H[i] <= 5, 0, 1)
  )
  single_gada_miaa$ia2h_cat[i]<- e
  ###coding for znt8
  f<-ifelse(is.na(single_gada_miaa$ZnT8[i]), 
            NA,
            ifelse(single_gada_miaa$ZnT8[i] <= 0.02, 0, 1)
  )
  single_gada_miaa$znt8_cat[i]<- f
}

#####combining both forms of gad65 tests
for(i in 1:nrow(single_gada_miaa)){
  test <- sum(single_gada_miaa$ica512_cat[i],single_gada_miaa$ia2h_cat[i],na.rm=TRUE)
  de <- ifelse(is.na(single_gada_miaa$ica512_cat[i]) & is.na(single_gada_miaa$ia2h_cat[i]),
               NA,
               ifelse(test>=1,1,0))
  single_gada_miaa$ica512_ia2_cat[i]<- de    
  test2 <- sum(single_gada_miaa$gad65_cat[i],single_gada_miaa$gad65h_cat[i],na.rm=TRUE)
  bc <- ifelse(is.na(single_gada_miaa$gad65_cat[i]) & is.na(single_gada_miaa$gad65h_cat[i]),
               1,
               ifelse(test2>=1,1,0))
  single_gada_miaa$GAD65_combo_cat[i]<- bc   
}  

# if all ab results missing, delete the record
misstwo <- single_gada_miaa[is.na(single_gada_miaa$mIAA) & is.na(single_gada_miaa$ica512_ia2_cat),]
single_gada_miaa <- single_gada_miaa[!is.na(single_gada_miaa$mIAA) | !is.na(single_gada_miaa$miaa_cat) | !is.na(single_gada_miaa$ica512_ia2_cat),]
# 36 records deleted

# check if anyone is missing GAD, ICA/IA2, or mIAA
missing <- single_gada_miaa[is.na(single_gada_miaa$GAD65_combo_cat) | is.na(single_gada_miaa$miaa_cat) | is.na(single_gada_miaa$ica512_ia2_cat),]
write.csv(missing,file="missing.csv") 

# get rid of visits with missing mIAA data
missmiaa <- single_gada_miaa[is.na(single_gada_miaa$GAD65_combo_cat) | is.na(single_gada_miaa$miaa_cat) | is.na(single_gada_miaa$ica512_ia2_cat),]
single_gada_miaa <- single_gada_miaa[!is.na(single_gada_miaa$GAD65_combo_cat) & !is.na(single_gada_miaa$miaa_cat) & !is.na(single_gada_miaa$ica512_ia2_cat),]
# 40 records deleted

# calculate number of positive ab at each visit
for(i in 1:nrow(single_gada_miaa)){
  single_gada_miaa$numpos[i] <- sum(single_gada_miaa$ica512_ia2_cat[i],single_gada_miaa$GAD65_combo_cat[i],single_gada_miaa$miaa_cat[i],single_gada_miaa$znt8_cat[i],na.rm=T)
}

# calculate z-scores for GAD and IA2
single_gada_miaa$gad65z <- (single_gada_miaa$GAD65 - mean(single_gada_miaa$GAD65,na.rm = TRUE))/sd(single_gada_miaa$GAD65,na.rm = TRUE)
single_gada_miaa$gad65hz <- (single_gada_miaa$GAD65H - mean(single_gada_miaa$GAD65H,na.rm = TRUE))/sd(single_gada_miaa$GAD65H,na.rm = TRUE)
single_gada_miaa$gadz <- ifelse(!is.na(single_gada_miaa$GAD65),single_gada_miaa$gad65z,single_gada_miaa$gad65hz)
single_gada_miaa$ica512z <- (single_gada_miaa$ICA512 - mean(single_gada_miaa$ICA512,na.rm = TRUE))/sd(single_gada_miaa$ICA512,na.rm = TRUE)
single_gada_miaa$ia2hz <- (single_gada_miaa$IA_2H - mean(single_gada_miaa$IA_2H,na.rm = TRUE))/sd(single_gada_miaa$IA_2H,na.rm = TRUE)
single_gada_miaa$ia2z <- ifelse(!is.na(single_gada_miaa$ICA512),single_gada_miaa$ica512z,single_gada_miaa$ia2hz)

# exclude people who do not have a proinsulin ratio 
sortds <- single_gada_miaa[order(single_gada_miaa$Random_ID,desc(single_gada_miaa$Pro_pep)),]
hasproins <- sortds %>% group_by(Random_ID) %>% filter(row_number(Random_ID) == 1)
a <- hasproins[is.na(hasproins$Pro_pep),]
a <- a[,1]
a$flaga <- rep(1,nrow(a))
# a contains a list of IDs of people missing proins
# people with flaga are missing proins and need to be excluded

exclu <- a[a$flaga==1,] 
misspro <- length(unique(exclu$Random_ID))
single_gada_miaa <- merge(single_gada_miaa,a,by="Random_ID",all.x=TRUE,all.y=TRUE)
single_gada_miaa$flaga[is.na(single_gada_miaa$flaga)] <- 0
single_gada_miaa <- single_gada_miaa[single_gada_miaa$flaga==0,]

# who is missing T1D data?
missingt1d <- single_gada_miaa[is.na(single_gada_miaa$T1D),]
# no one is missing t1d status

# group status at each visit - but people can change groups
for(i in 1:nrow(single_gada_miaa)) {
  single_gada_miaa$group[i] <- ifelse(single_gada_miaa$T1D[i]=="Yes","T1D",
                             ifelse(single_gada_miaa$numpos[i]==0,"Negative",
                                    ifelse(single_gada_miaa$numpos[i]==1,"Single ab positive","Multiple ab positive")))
}
for(i in 1:nrow(single_gada_miaa)) {
  single_gada_miaa$newgroup[i] <- ifelse(single_gada_miaa$T1D[i]=="Yes","T1D",
                             ifelse(single_gada_miaa$numpos[i]==0,"Negative",
                                    ifelse(single_gada_miaa$numpos[i]==1 & single_gada_miaa$miaa_cat[i]==1,"Single mIAA positive",
                                           ifelse(single_gada_miaa$numpos[i]==1 & single_gada_miaa$GAD65_combo_cat[i]==1,"Single GAD positive",
                                                  ifelse(single_gada_miaa$numpos[i]==1,"Other single ab positive","Multiple ab positive")))))
}

#-------------------------------------------------------------------------------
# Tim's group status code.
# Separate table for diagnosis.
diagnosis.table <- single_gada_miaa[,c("Random_ID","Draw_Date","group")]
# save a copy of this data to merge back with the final dx
allresults <- diagnosis.table
# Count diagnoses per subject
counts <- plyr::count(diagnosis.table,c("Random_ID","group"))
# Find participants with only one single ab+
single <- counts[counts$group == "Single ab positive",]
single <- single[single$freq < 2,"Random_ID"]
# Convert group to numeric
diagnosis.table$group <- 
  plyr::revalue(diagnosis.table$group,c("Negative" = 0,"Single ab positive" = 1,
                                        "Multiple ab positive" = 2,"T1D" = 3))
diagnosis.table$group <- as.numeric(diagnosis.table$group)
# Aggregate diagnosis table taking the maximum value.
diagnosis.table <- aggregate(group ~ Random_ID, data = diagnosis.table, max)
# Find the single ab+ who were incorrectly categorized, set back to 0.
recat <- single[
  which(single %in% diagnosis.table[diagnosis.table$group == 1,"Random_ID"])]
diagnosis.table[diagnosis.table$Random_ID %in% recat,"group"] <- 0
colnames(diagnosis.table) <- c("Random_ID","final.group")
# make categorical variable for final group
diagnosis.table$final.group[diagnosis.table$final.group==0] <- "Negative"
diagnosis.table$final.group[diagnosis.table$final.group==1] <- "Single ab positive"
diagnosis.table$final.group[diagnosis.table$final.group==2] <- "Multiple ab positive"
diagnosis.table$final.group[diagnosis.table$final.group==3] <- "T1D"
# merge final group category with all results
allresults <- merge(allresults,diagnosis.table,by="Random_ID")
single_gada_miaa <- merge(single_gada_miaa,diagnosis.table,by = "Random_ID")
# create new variable for group with T1D and multiple ab combined
#single_gada_miaa$group2 <- ifelse(single_gada_miaa$final.group=="T1D" | single_gada_miaa$final.group=="Multiple ab positive","T1D/multiple ab pos",single_gada_miaa$final.group)

# of those who were previously classified as single antibody positive, describe # of single ab positive visits
singlecheck <- single_gada_miaa[single_gada_miaa$final.group=="Single ab positive",
                       c("Random_ID","Draw_Date","final.group","newgroup","ica512_ia2_cat","GAD65_combo_cat","znt8_cat","miaa_cat")]
# Count visits by ab status for each person classified as single ab positive
singlecounts <- plyr::count(singlecheck,c("Random_ID","newgroup"))
# reshape the count dataframe
singlecount_wide <- reshape(singlecounts,timevar="newgroup",idvar="Random_ID",direction="wide")
singlecount_wide$newcat[singlecount_wide$`freq.Single GAD positive`>=2 & is.na(singlecount_wide$`freq.Other single ab positive`)
                        &is.na(singlecount_wide$`freq.Single mIAA positive`)] <- "Single GAD positive"
singlecount_wide$newcat[is.na(singlecount_wide$`freq.Single GAD positive`) & is.na(singlecount_wide$`freq.Single mIAA positive`)] <- "Other single ab positive"
singlecount_wide$newcat[singlecount_wide$`freq.Single mIAA positive`>=2 & is.na(singlecount_wide$`freq.Other single ab positive`)
                        &is.na(singlecount_wide$`freq.Single GAD positive`)] <- "Single mIAA positive"

forandrea <- merge(singlecheck,singlecount_wide,by="Random_ID")

#write.csv(forandrea,"S:\\Shared Projects\\Laura\\BDC\\Projects\\Andrea Steck\\Trialnet progressors analysis\\Fourth analysis - demethylated insulin #with updated dataset\\Data\\checking_single_ab.csv")

# FOR THIS ANALYSIS, KEEP ONLY THE SINGLE GADA AND SINGLE MIAA POSITVE PEOPLE
single_gada_miaa <- merge(single_gada_miaa,singlecount_wide,by="Random_ID")
single_gada_miaa <- single_gada_miaa[single_gada_miaa$newcat %in% c("Single GAD positive","Single mIAA positive"),]

#-------------------------------------------------------------------------------


# find the first draw date
single_gada_miaa <- single_gada_miaa[order(single_gada_miaa$Random_ID,single_gada_miaa$Draw_Date),]
temp <- aggregate(single_gada_miaa$Draw_Date,by=list(single_gada_miaa$Random_ID),min)
names(temp) <- c("Random_ID","first_date")
temp$first_date <- as.Date(temp$first_date)
single_gada_miaa <- merge(single_gada_miaa,temp,by = "Random_ID")
single_gada_miaa$Draw_Date <- as.Date(single_gada_miaa$Draw_Date)
single_gada_miaa$time <- single_gada_miaa$Draw_Date - single_gada_miaa$first_date
single_gada_miaa$years <- single_gada_miaa$time/365.25

# check range of years for non-missing pro-pep
a <- single_gada_miaa[!is.na(single_gada_miaa$Pro_pep),]
hist(as.numeric(a$years))


# spaghetti plots of outcomes
#qplot(years,Pro_pep,data=single_gada_miaa,group=Random_ID,color=group,geom="line")
#qplot(years,U_MplusU_Ratio_Zen,data=single_gada_miaa,group=Random_ID,color=group,geom="line")

# now need to determine which INS ratio/proinsulin ratio to use for cross-sectional analyses
# T1D: results closest to T1D onset 
single_gada_miaa$Date_of_Diagnosis <- as.Date(mdy(single_gada_miaa$Date_of_Diagnosis))
single_gada_miaa$timec[single_gada_miaa$final.group=="T1D"] <- abs(single_gada_miaa$Draw_Date[single_gada_miaa$final.group=="T1D"] - single_gada_miaa$Date_of_Diagnosis[single_gada_miaa$final.group=="T1D"])

# create new variable for IGTT
single_gada_miaa$igtt1[single_gada_miaa$GLU0>100 & single_gada_miaa$GLU0<126] <- 1
single_gada_miaa$igtt2[single_gada_miaa$GLU120>= 140 & single_gada_miaa$GLU120<200] <- 1
single_gada_miaa$igtt3[single_gada_miaa$GLU30>200 | single_gada_miaa$GLU60>200 | single_gada_miaa$GLU90>200] <- 1
single_gada_miaa$igtt[single_gada_miaa$igtt1==1 | single_gada_miaa$igtt2==1 | single_gada_miaa$igtt3==1] <- 1
# now need to take maximum of IGTT to see if they ever had IGTT
igttcheck <- single_gada_miaa[c("Random_ID","igtt")]
igttcheck$igtt[is.na(igttcheck$igtt)] <- 0
igttever <- aggregate(igttcheck$igtt,by=list(igttcheck$Random_ID),max)
names(igttever) <- c("Random_ID","igttever")
single_gada_miaa <- merge(single_gada_miaa,igttever,by="Random_ID")
single_gada_miaa$igttever[is.na(single_gada_miaa$igttever)] <- 0
single_gada_miaa$age <- floor(single_gada_miaa$DOB %--% single_gada_miaa$Draw_Date / dyears(1))

# calculate follow up time
firstvisit <- single_gada_miaa[order(single_gada_miaa$Random_ID,single_gada_miaa$Draw_Date),]
firstvisit <- firstvisit[,c("Random_ID","Draw_Date","Date_of_Diagnosis")]
firstvisit <- firstvisit %>% group_by(Random_ID) %>% filter(row_number(Random_ID) == 1)
names(firstvisit) <- c("Random_ID","firstdate","Date_of_Diagnosis")
lastvisit <- single_gada_miaa[order(single_gada_miaa$Random_ID,desc(single_gada_miaa$Draw_Date)),]
lastvisit <- lastvisit[,c("Random_ID","Draw_Date")]
names(lastvisit) <- c("Random_ID","lastdate")
lastvisit <- lastvisit %>% group_by(Random_ID) %>% filter(row_number(Random_ID) == 1)
firstlast <- merge(firstvisit,lastvisit,by="Random_ID",all=TRUE)
firstlast$lastdate <- ifelse(!is.na(firstlast$Date_of_Diagnosis) & firstlast$Date_of_Diagnosis>firstlast$lastdate,
                             firstlast$Date_of_Diagnosis,firstlast$lastdate)
firstlast$lastdate <- as.Date(firstlast$lastdate,origin = "1970-01-01")
firstlast$followup <-  as.numeric(firstlast$lastdate - firstlast$firstdate) / 365.25
single_gada_miaa <- merge(single_gada_miaa,firstlast,by="Random_ID",all=TRUE)

futab <- tapply(single_gada_miaa$followup,single_gada_miaa$final.group,summary)

# labels
label(single_gada_miaa$Pro_pep)='Proinsulin ratio'
label(single_gada_miaa$Gender)='Gender'
label(single_gada_miaa$DPT1Risk)="DPT1 Risk Score"
label(single_gada_miaa$DR3)="DR3"
label(single_gada_miaa$DR4)="DR4"
label(single_gada_miaa$DR3DR4)="DR3DR4"
label(single_gada_miaa$BMI)="BMI"
label(single_gada_miaa$followup)="Follow-up time (yrs)"

# create dataset with earliest proinsulin ratio
earliest_pro <- single_gada_miaa[!is.na(single_gada_miaa$Pro_pep),]
earliest_pro <- earliest_pro[order(earliest_pro$Random_ID,earliest_pro$Draw_Date),]
earliest_pro <- earliest_pro %>% group_by(Random_ID) %>% filter(row_number() == 1)
earliest_pro$age <- floor(earliest_pro$DOB %--% earliest_pro$Draw_Date / dyears(1))

# create labeled factor for final.group
#check <- single_gada_miaa %>% distinct(Random_ID, .keep_all=TRUE)
#table(check$final.group)
#single_gada_miaa$final.group <- as.factor(single_gada_miaa$final.group)
#check <- single_gada_miaa %>% distinct(Random_ID, .keep_all=TRUE)
#table(check$final.group)
#levels(single_gada_miaa$final.group) = c("Negative","Single ab positive","Multiple ab positive","T1D")
#check <- single_gada_miaa %>% distinct(Random_ID, .keep_all=TRUE)
#table(check$final.group)

# pull out T1D - if they have T1D at any time point, they are in the T1D group
#t1d <- single_gada_miaa[single_gada_miaa$T1D == "Yes",]
#t1d <- t1d[order(t1d$Random_ID, t1d$Draw_Date),]
# pull out nonT1D
nont1d <- single_gada_miaa[single_gada_miaa$T1D != "Yes",]
nont1d <- nont1d[order(nont1d$Random_ID, nont1d$Draw_Date),]
ngada <- length(unique(nont1d[nont1d$newcat=="Single GAD positive",]$Random_ID))
nmiaa <- length(unique(nont1d[nont1d$newcat=="Single mIAA positive",]$Random_ID))
#nmult <- length(unique(nont1d[nont1d$final.group=="Multiple ab positive",]$Random_ID))
# total of 28 single GADA and 14 single mIAA
# pull out records with non-missing Pro_pep
#t1d_select_pro <- t1d[!is.na(t1d$Pro_pep),]
#temp <-  aggregate(t1d_select_pro$timec,by=list(t1d_select_pro$Random_ID),min)
#names(temp) <- c("Random_ID","time_from_diagnosis")
#t1d_select_pro <- merge(t1d_select_pro,temp,by = "Random_ID")
#t1d_select_pro <- t1d_select_pro[t1d_select_pro$time_from_diagnosis==t1d_select_pro$timec,]

# now if the proinsulin value closest to dx is not within 3 months, need to use group at the time of proinsulin
#for(i in 1:nrow(t1d_select_pro)) {
#  t1d_select_pro$final.group[i] <- ifelse(t1d_select_pro$time_from_diagnosis[i]<=90, "T1D",    
#                                   ifelse(t1d_select_pro$numpos[i]==0,"Negative",
#                                    ifelse(t1d_select_pro$numpos[i]==1,"Single ab positive","Multiple ab positive")))
#}

# for non-T1D group, use the most recent proinsulin value that corresponds to the final ab group
# if there isn't a proinsulin value that corresponds to final ab group, use the most recent proinsulin value
# available and reclassify the person into the ab group corresponding to that visit
nont1D_match <- nont1d[!is.na(nont1d$Pro_pep),]
#check <- nont1d_select_pro %>% distinct(Random_ID, .keep_all=TRUE)
#nont1d_select_pro$flag[nont1d_select_pro$final.group == nont1d_select_pro$group] <- 1
# create another df for people who are flagged who have a propep that matches their final group
#nont1D_match <- nont1d_select_pro[nont1d_select_pro$flag==1,]
# keep the most recent proinsulin for those people
nont1D_match <- nont1D_match[order(nont1D_match$Random_ID,desc(nont1D_match$Draw_Date)),]
nont1D_match <- nont1D_match %>% group_by(Random_ID) %>% filter(row_number(Random_ID) == 1)
# go back to original dataset and take out people who had a match
# this leaves only people who do not have a proinsulin value drawn while they were in their final.group category
#x <- as.data.frame(nont1D_match$Random_ID)
#x$flagx <- rep(1,length(x))
#names(x) <- c("Random_ID","flagx")
#nont1d_select_pro <- merge(nont1d_select_pro,x,by="Random_ID",all.x = TRUE,all.y=TRUE)
#nont1d_nomatch <- nont1d_select_pro[is.na(nont1d_select_pro$flagx),]
# keep the most recent proinsulin value
#nont1d_nomatch <- nont1d_nomatch[order(nont1d_nomatch$Random_ID,desc(nont1d_nomatch$Draw_Date)),]
#nont1d_nomatch <- nont1d_nomatch %>% group_by(Random_ID) %>% filter(row_number(Random_ID) == 1)
# now reclassify final.group to be equal to current group
#nont1d_nomatch$final.group <- nont1d_nomatch$group
# combine those with match and those without
nont1d_select_pro <- nont1D_match
# people in match group: 68
# people in no match group: 17
# people in final dataset: 85

# combine the results for the pro-pep analysis
#t1d_select_pro$most_recent_draw <- NA
nont1d_select_pro$time_from_diagnosis <- NA
allpro <- nont1d_select_pro
allpro$age <- floor(allpro$DOB %--% allpro$Draw_Date / dyears(1))
label(allpro$Pro_pep)="Proinsulin ratio"

# correct group2 variable in allpro
#allpro$group2 <- ifelse(allpro$final.group=="T1D" | allpro$final.group=="Multiple ab positive","T1D/multiple ab pos",allpro$final.group)


# for table 1, need the draw date corresponding to the proins used in cross sectional analyses
datepro <- allpro[c("Random_ID","Draw_Date")]
names(datepro) <- c("Random_ID","Date_proins")
getdates <- datepro
# keep the latest date to find A1c and age
getdates$tab1_date <- pmax(getdates$Date_proins,na.rm=TRUE)
getdates$tab1flag <- rep(1,nrow(getdates))
getdates$Draw_Date <- getdates$tab1_date
tab1data <- merge(single_gada_miaa,getdates,by=c("Random_ID","Draw_Date"))
tab1data$age <- floor(tab1data$DOB %--% tab1data$Draw_Date / dyears(1))
label(tab1data$Pro_pep)="Proinsulin ratio"


# need to update final.group variable using what is in allpro
#correct_group <- subset(allpro, select =c(Random_ID,final.group))
#tab1data <- subset(tab1data, select = -c(final.group))
#tab1data <- merge(tab1data,correct_group,by="Random_ID",all=TRUE)

# ++++++++++++++++++++++++++++
# flattenCorrMatrix
# ++++++++++++++++++++++++++++
# cormat : matrix of the correlation coefficients
# pmat : matrix of the correlation p-values
flattenCorrMatrix <- function(cormat, pmat) {
  ut <- upper.tri(cormat)
  data.frame(
    row = rownames(cormat)[row(cormat)[ut]],
    column = rownames(cormat)[col(cormat)[ut]],
    cor  =(cormat)[ut],
    p = pmat[ut]
    )
}



# compare proinsulin ratios between T1D, Ab-, single Ab+, multiple Ab+

# unadjusted comparisons
tab_pro <- final_table(allpro,'Pro_pep',group=as.factor(allpro$newcat),margin=2,single=0,ron=2)

# adjust for age, sex, and BMI
allpro$l_Pro_pep=log(allpro$Pro_pep)
# lm.pro = with(allpro, lm(Pro_pep ~ as.factor(final.group))) 
# summary(lm.pro)
# anova(lm.pro)
# plot(lm.pro)
# lm.pro <- aov(lm.pro)
# summary(lm.pro)
# anova(lm.pro)
# TukeyHSD(lm.pro)
# log transformed
lm.pro = with(allpro, lm(Pro_pep ~ as.factor(newcat) + age + Gender + BMI)) 
summary(lm.pro)
anova(lm.pro)
lm.pro <- aov(lm.pro)
summary(lm.pro)
anova(lm.pro)
pval_newcat <- anova(lm.pro)[1,5]
diffs_newcat <- TukeyHSD(lm.pro, which = "as.factor(newcat)")


# do proinsulin ratios correlate with biochemical and ECL antibody levels
corr_pro <- rcorr(as.matrix(allpro[, c( "Pro_pep","gadz","ia2z","mIAA","ZnT8","ICA",
                                        "ECL_GADA","ECL_IA2A","ECL_IAA")]),type = "spearman")
corr_pro <- flattenCorrMatrix(corr_pro$r, corr_pro$P)
corr_pro <- corr_pro[corr_pro$row=="Pro_pep",]

# correlations in single mIAA positive, just mIAA and ECL_IAA
corr_pro_miaa <- rcorr(as.matrix(allpro[allpro$newcat=="Single mIAA positive", c( "Pro_pep","mIAA","ECL_IAA")]),type = "spearman")
corr_pro_miaa <- flattenCorrMatrix(corr_pro_miaa$r, corr_pro_miaa$P)
corr_pro_miaa <- corr_pro_miaa[corr_pro_miaa$row=="Pro_pep",]

# correlations in single GADA positive, just GADA and ECL_GADA
corr_pro_gad <- rcorr(as.matrix(allpro[allpro$newcat=="Single GAD positive", c( "Pro_pep","gadz","ECL_GADA")]),type = "spearman")
corr_pro_gad <- flattenCorrMatrix(corr_pro_gad$r, corr_pro_gad$P)
corr_pro_gad <- corr_pro_gad[corr_pro_gad$row=="Pro_pep",]


# partial correlations adjusted for age, sex, gender
#blah <- na.omit(remove_all_labels(allpro))
#corr_pro_part <- pcor.test(x=blah$Pro_pep,y=blah[,c("gadz","ia2z","mIAA","ZnT8","ICA","ECL_GADA","ECL_IA2A","ECL_IAA")],
#                           z=blah[,c("age","BMI")])

# table 1 - include age at last visit, gender, race/eth, DPT1 score, HLA, A1c
label(tab1data$age)="Age (yrs)"
label(tab1data$HLA)="HLA"
label(tab1data$HbA1c)="HbA1c (%)"
label(tab1data$race_eth)="Race/ethnicity"
tab1data$DR3DR4 <- as.factor(tab1data$DR3DR4)
label(tab1data$DR3DR4)="DR3 DR4"

t1 <- final_table(tab1data,c("age","Gender","race_eth","DPT1Risk","DR3","DR4","DR3DR4","HLA","HbA1c","BMI","followup"),group=as.factor(tab1data$newcat),margin=2,single=0,ron=1)

# does earliest proins value predict T1D or IGTT?
#log1 <- glm(T1D~Pro_pep , earliest_pro, family='binomial')
#log_t1d <- mod_tablog(log1)
#log_t1d$Predictor<-c("Proinsulin ratio")
#log2 <- glm(igttever~Pro_pep , earliest_pro, family='binomial')
#log_igtt <- mod_tablog(log2)
#log_igtt$Predictor<-c("Proinsulin ratio")

#log1_adj <- glm(T1D~Pro_pep + age + as.factor(Gender) + BMI , earliest_pro, family='binomial')
#log_t1d_adj <- mod_tablog(log1_adj)
#log_t1d_adj$Predictor<-c("Proinsulin ratio","Age","Male sex","BMI")
#log2_adj <- glm(igttever~Pro_pep + age + Gender + BMI, earliest_pro, family='binomial')
#log_igtt_adj <- mod_tablog(log2_adj)
#log_igtt_adj$Predictor<-c("Proinsulin ratio","Age","Male sex","BMI")

# longitudinal analysis
long <- single_gada_miaa[!is.na(single_gada_miaa$time),]
mix <- lme(Pro_pep~ time + as.factor(newcat) + time:as.factor(newcat), random = ~1 + time | Random_ID, data=long, na.action = na.omit)
# type III tests
mixt <- anova.lme(mix,type="marginal")

# longitudinal analysis
mix_adj <- lme(Pro_pep~ time + as.factor(newcat) + time:as.factor(newcat) + age + Gender + BMI, 
               random = ~1 + time | Random_ID, data=long, na.action = na.omit)
# type III tests
mixt_adj <- anova.lme(mix_adj,type="marginal")

```

## Background

The purpose of this analysis is to test whether proinsulin ratios differ between 2 groups of TrialNet subjects: single GADA positive and single mIAA positive.  The correlation of proinsulin ratios with biochemical and ECL antibody levels was also examined.

## Methods

There were `r obslong` records in the initial longitudinal dataset, and `r nppt` participants in the participant database.

The following data manipulation and cleaning steps were performed:

1) Fasting C-peptide in pM/L was calculated as fasting C-peptide*331.09.
2) Proinsulin ratios were calculated as proinsulin (pM/L) / C-peptide (pM/L) *100
3) There were `r nrow(misstwo)` records deleted that were missing both mIAA and ICA512/IA2H results.
4) There were an additional `r nrow(missmiaa)` records deleted that were missing just mIAA.
5) Antibody results were classifed as positive/negative using the following cutoff values:
+ mIAA, 0.01
+ GAD65, 0.032
+ GAD65H, 20
+ ICA512, 0.049
+ IA_2H, 5
+ ZnT8, 0.02
6) The categorical variables (positive/negative) for GAD65 and GAD65H were combined into one variable, as were the categorical variables for ICA512 and IA_2H.
7) The number of positive antibodies at each visit was calculated as the sum of ICA512/IA2H, GAD65/GAD65H, mIAA, and ZnT8.
8) For analyses using the continuous value of the antibodies, GAD65 and GAD65H were combined by calculating the Z-score of each variable and then combining the Z-scores into one variable.  The same procedure was followed for ICA512/IA2H.
9) There were `r misspro` subjects who did not have any proinsulin ratio results who were excluded from analysis.   
10) If a participant did not have a diagnosis of T1D and was never multiple antibody positive at any visit, but was single antibody positive at 2 or more visits, they were classified as single antibody positive.
11) Those who were single antibody positive were further classified as follows:
+ Single mIAA positive (at least 2 visits with only mIAA positivity, and no visits single antibody positive for another antibody); N=`r nmiaa`
+ Single GADA positive (at least 2 visits with only GADA positivity, and no visits single antibody positive for another antibody); N=`r ngada`
+ There was one participant who had mixed single antibody positivity; this participant was excluded from analysis. 
12) For this analysis, participants who had T1D or who were multiple antibody positive or antibody negative were excluded from analysis.
13) For cross-sectional analyses, the most recent proinsulin ratios was used.
14) Follow-up time was calculated as the years elapsed between the first draw date and the last draw date or the date of diagnosis, whichever was later.

The distribution of all variables was asssessed prior to analysis.  Group comparisons were done using ANOVA or the Kruskal-Wallis test for continuous variables and the chi-square or Fisher's exact test for categorical variables.  A random coefficients mixed model was used to test whether the groups had different trajectories of proinsulin over time.

## Results

Table 1a presents descriptive statistics for demographic and clinical variables by group.

```{r, results='asis',tidy=TRUE, echo=FALSE}
kable(t1,caption="Table 1a.  Descriptive statistics.  Asterisks indicate that Fisher's exact test was used instead of the chi-square test due to small expected cell value.")
```
\

```{r, results='asis',tidy=TRUE, echo=FALSE}
kable(futab,caption="Table 1b.  Descriptive statistics for follow up time by group.")
```
\

Table 2a shows the results of the group comparisons for proinsulin ratios.

```{r, results='asis',tidy=TRUE, echo=FALSE}
kable(tab_pro,caption="Table 2a.  Proinsulin ratios.")
```
\

After adjusting for age, sex, and BMI, the groups were not significantly different (p=`r pval_newcat`).  Table 3 shows the differences between all pairwise comparisons of groups, using Tukey's HSD test.

```{r, results='asis',tidy=TRUE, echo=FALSE}
kable(diffs_newcat$`as.factor(newcat)`,caption="Table 3.  Pairwise comparisons of least-squares mean proinsulin ratio, adjusted for age, sex, and BMI.")
```
\

Table 4a shows the correlations of proinsulin ratios with continuous antibody levels at the timepoint selected for cross-sectional analysis, in both single mIAA positive and single GADA positive subjects.

```{r, results='asis',tidy=TRUE, echo=FALSE}
kable(corr_pro,caption="Table 4a.  Correlation of proinsulin ratios with antibody levels, in both single mIAA positive and single GADA positive subjects.")
```
\

Table 4b shows the correlations of proinsulin ratios with continuous antibody levels at the timepoint selected for cross-sectional analysis, in single mIAA positive subjects only.

```{r, results='asis',tidy=TRUE, echo=FALSE}
kable(corr_pro_miaa,caption="Table 4b.  Correlation of proinsulin ratios with antibody levels, in single mIAA positive subjects only.")
```
\

Table 4c shows the correlations of proinsulin ratios with continuous antibody levels at the timepoint selected for cross-sectional analysis, in single GADA positive subjects only.

```{r, results='asis',tidy=TRUE, echo=FALSE}
kable(corr_pro_gad,caption="Table 4c.  Correlation of proinsulin ratios with antibody levels, in single GADA positive subjects only.")
```
\

Table 7 shows the results of the mixed model testing whether the groups differ in the trajectories of proinsulin ratios over time.  The trajectories in the two groups were not significantly different (p=`r mixt[4,4]`).

```{r, results='asis',tidy=TRUE, echo=FALSE}
kable(mixt,caption="Table 7.  ANOVA table for mixed effects model.")
```
\

Table 8 shows the results of the mixed model testing whether the groups differ in the trajectories of proinsulin ratios over time, after adjusting for sex, age, and BMI.  The trajectories in the four groups were not significantly different (p=`r mixt_adj[7,4]`).

```{r, results='asis',tidy=TRUE, echo=FALSE}
kable(mixt_adj,caption="Table 8.  ANOVA table for mixed effects model, adjusted for sex, age, and BMI.")
```
\

