---
title: "TrialNet INS ratio analyses"
author: "Laura Pyle"
date: "November 15, 2018"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

#library(sqldf)
library(dplyr)
library(lubridate)
library(graphics)
library(ggplot2)
library(Hmisc)
library(Table1)
library(knitr)

# read in the data
longitudinal_visit_info <- read.csv("\\\\ucdenver.pvt\\som-home\\pylell\\BDC\\Steck\\Trialnet progressors analysis\\Fourth analysis - demethylated insulin with updated dataset\\Data\\longitudinal 091818.csv")
ParticipantInfo_risk <- read.csv("\\\\ucdenver.pvt\\som-home\\pylell\\BDC\\Steck\\Trialnet progressors analysis\\Fourth analysis - demethylated insulin with updated dataset\\Data\\Participant Info risk 091818.csv",na.strings = c("NA"," ","","      "))

# get rid of duplicates
ParticipantInfo_risk <- unique(ParticipantInfo_risk)
# no duplicates

obslong <- nrow(longitudinal_visit_info)
nppt <- nrow(ParticipantInfo_risk)

# merge participant info and longitudinal data
keep_ppt <- data.frame(ParticipantInfo_risk$TrialNet,
                       ParticipantInfo_risk$T1D,
                       ParticipantInfo_risk$Date.of.Diagnosis,
                       ParticipantInfo_risk$DOB, ParticipantInfo_risk$Gender, ParticipantInfo_risk$Race, ParticipantInfo_risk$Ethnicity,
                       ParticipantInfo_risk$DPT1.Risk.Score,ParticipantInfo_risk$HLA, ParticipantInfo_risk$DR3,ParticipantInfo_risk$DR4) 
colnames(keep_ppt) <- c("Random_ID","T1D","Date_of_Diagnosis","DOB","Gender",
                        "Race","Ethnicity","DPT1Risk","HLA","DR3","DR4")
keep_long <- data.frame(longitudinal_visit_info$PID,
                        longitudinal_visit_info$Draw.Date,
                        longitudinal_visit_info$GAD65,
                        longitudinal_visit_info$GAD65H,
                        longitudinal_visit_info$ICA512,
                        longitudinal_visit_info$IA2H,
                        longitudinal_visit_info$mIAA,
                        longitudinal_visit_info$ZnT8,
                        longitudinal_visit_info$ICA,
                        longitudinal_visit_info$U.per.MplusU.Ratio.Zen,
                        longitudinal_visit_info$ECL_GADA,
                        longitudinal_visit_info$ECL_IA.2A,
                        longitudinal_visit_info$ECL_IAA,
                        longitudinal_visit_info$Proinsulin_pM.L,
                        longitudinal_visit_info$PEP0,
                        longitudinal_visit_info$HbA1c)
colnames(keep_long) <- c("Random_ID","Draw_Date","GAD65","GAD65H","ICA512","IA_2H","mIAA","ZnT8","ICA","U_MplusU_Ratio_Zen","ECL_GADA","ECL_IA2A","ECL_IAA","Proinsulin_pM_L","PEP0","HbA1c")

# convert draw date to date
#keep_long$Draw_Date <- mdy_hms(as.character(keep_long$Draw_Date))
#keep_long$Draw_Date <- as.POSIXlt(as.character(keep_long$Draw_Date),format="%m%d%Y %H%M")
keep_long$Draw_Date <- mdy_hm(as.character(keep_long$Draw_Date))
keep_ppt$DOB <- mdy(as.character(keep_ppt$DOB))

# calculate fasting Cpeptide in pM/L
keep_long$PEP0_pM_L <- keep_long$PEP0*331.09

# calculate proinsulin/Cpep ratio
keep_long$Pro_pep <- keep_long$Proinsulin_pM_L/keep_long$PEP0_pM_L*100

# merge  datasets
alldata <- merge(keep_ppt,keep_long,by="Random_ID",all.x=F,all.y=F)
N <- length(unique(alldata$Random_ID))

# fix gender coding female Female   male   Male 
alldata$Gender[alldata$Gender=='female'] <- "Female"
alldata$Gender[alldata$Gender=='male'] <- "Male"
alldata$Gender <- droplevels(alldata$Gender)

# new variable for DR3/DR4
alldata$DR3DR4[alldata$DR3=="PRESENT" & alldata$DR4=="PRESENT"] <- "PRESENT"
alldata$DR3DR4[!(alldata$DR3=="PRESENT" & alldata$DR4=="PRESENT")] <- "ABSENT"
alldata$DR3DR4[is.na(alldata$DR3) & is.na(alldata$DR4)] <- NA


# combine race/ethnicity
alldata$race_eth <- NA
alldata$race_eth[(alldata$Race=="White" & alldata$Ethnicity=="Non-Hispanic origin")]  <- "NHW"
alldata$race_eth[(alldata$Race=="White" & alldata$Ethnicity=="Non-Hispanic White")] <- "NHW"
alldata$race_eth[(alldata$Race=="White" & alldata$Ethnicity == "Not Hispanic or Latino")] <- "NHW"
alldata$race_eth[alldata$Ethnicity=="Hispanic" | alldata$Ethnicity=="Hispanic or Latino" | alldata$Ethnicity=="Hispanic/Latino/Spanish origin"] <- "Hispanic"
alldata$race_eth[is.na(alldata$race_eth)] <- "Other"
alldata$race_eth <- as.factor(alldata$race_eth)


# labels
label(alldata$U_MplusU_Ratio_Zen)='INS ratio'
label(alldata$Pro_pep)='Proinsulin ratio'
label(alldata$Gender)='Gender'
label(alldata$DPT1Risk)="DPT1 Risk Score"
label(alldata$DR3)="DR3"
label(alldata$DR4)="DR4"
label(alldata$DR3DR4)="DR3/DR4"

# first code each antibody as positive negative
for(i in 1:nrow(alldata)){
  ###coding for miaa
  a <-ifelse(is.na(alldata$mIAA[i]), 
             NA,
             ifelse(alldata$mIAA[i] <= 0.01, 0, 1)
  )
  alldata$miaa_cat[i]<- a
  ###coding for first form of gad65
  b <-ifelse(is.na(alldata$GAD65[i]), 
             NA,
             ifelse(alldata$GAD65[i] <= 0.032, 0, 1)
  )
  alldata$gad65_cat[i]<- b
  ###coding for second form of gad65
  c<-ifelse(is.na(alldata$GAD65H[i]), 
            NA,
            ifelse(alldata$GAD65H[i] <= 20, 0, 1)
  )
  alldata$gad65h_cat[i]<- c
  
  ###coding for first form of ica512/ia2
  d<-ifelse(is.na(alldata$ICA512[i]), 
            NA,
            ifelse(alldata$ICA512[i] <= 0.049, 0, 1)
  )
  alldata$ica512_cat[i]<- d
  ###coding for second form of ica512/ia2
  e<-ifelse(is.na(alldata$IA_2H[i]), 
            NA,
            ifelse(alldata$IA_2H[i] <= 5, 0, 1)
  )
  alldata$ia2h_cat[i]<- e
  ###coding for znt8
  f<-ifelse(is.na(alldata$ZnT8[i]), 
            NA,
            ifelse(alldata$ZnT8[i] <= 0.02, 0, 1)
  )
  alldata$znt8_cat[i]<- f
}

#####combining both forms of gad65 tests
for(i in 1:nrow(alldata)){
  test <- sum(alldata$ica512_cat[i],alldata$ia2h_cat[i],na.rm=TRUE)
  de <- ifelse(is.na(alldata$ica512_cat[i]) & is.na(alldata$ia2h_cat[i]),
               NA,
               ifelse(test>=1,1,0))
  alldata$ica512_ia2_cat[i]<- de    
  test2 <- sum(alldata$gad65_cat[i],alldata$gad65h_cat[i],na.rm=TRUE)
  bc <- ifelse(is.na(alldata$gad65_cat[i]) & is.na(alldata$gad65h_cat[i]),
               1,
               ifelse(test2>=1,1,0))
  alldata$GAD65_combo_cat[i]<- bc   
}  

# if all ab results missing, delete the record
misstwo <- alldata[is.na(alldata$mIAA) & is.na(alldata$ica512_ia2_cat),]
alldata <- alldata[!is.na(alldata$mIAA) | !is.na(alldata$miaa_cat) | !is.na(alldata$ica512_ia2_cat),]
# 36 records deleted

# check if anyone is missing GAD, ICA/IA2, or mIAA
missing <- alldata[is.na(alldata$GAD65_combo_cat) | is.na(alldata$miaa_cat) | is.na(alldata$ica512_ia2_cat),]
write.csv(missing,file="missing.csv") 

# get rid of visits with missing mIAA data
missmiaa <- alldata[is.na(alldata$GAD65_combo_cat) | is.na(alldata$miaa_cat) | is.na(alldata$ica512_ia2_cat),]
alldata <- alldata[!is.na(alldata$GAD65_combo_cat) & !is.na(alldata$miaa_cat) & !is.na(alldata$ica512_ia2_cat),]
# 40 records deleted

# calculate number of positive ab at each visit
for(i in 1:nrow(alldata)){
  alldata$numpos[i] <- sum(alldata$ica512_ia2_cat[i],alldata$GAD65_combo_cat[i],alldata$miaa_cat[i],alldata$znt8_cat[i],na.rm=T)
}

# calculate z-scores for GAD and IA2
alldata$gad65z <- (alldata$GAD65 - mean(alldata$GAD65,na.rm = TRUE))/sd(alldata$GAD65,na.rm = TRUE)
alldata$gad65hz <- (alldata$GAD65H - mean(alldata$GAD65H,na.rm = TRUE))/sd(alldata$GAD65H,na.rm = TRUE)
alldata$gadz <- ifelse(!is.na(alldata$GAD65),alldata$gad65z,alldata$gad65hz)
alldata$ica512z <- (alldata$ICA512 - mean(alldata$ICA512,na.rm = TRUE))/sd(alldata$ICA512,na.rm = TRUE)
alldata$ia2hz <- (alldata$IA_2H - mean(alldata$IA_2H,na.rm = TRUE))/sd(alldata$IA_2H,na.rm = TRUE)
alldata$ia2z <- ifelse(!is.na(alldata$ICA512),alldata$ica512z,alldata$ia2hz)

# exclude people who do not have a proinsulin ratio or INS ratio
sortds <- alldata[order(alldata$Random_ID,desc(alldata$Pro_pep)),]
hasproins <- sortds %>% group_by(Random_ID) %>% filter(row_number(Random_ID) == 1)
a <- hasproins[is.na(hasproins$Pro_pep),]
a <- a[,1]
a$flaga <- rep(1,nrow(a))
# a contains a list of IDs of people missing proins
sortds1 <- alldata[order(alldata$Random_ID,desc(alldata$U_MplusU_Ratio_Zen)),]
hasins <- sortds1 %>% group_by(Random_ID) %>% filter(row_number(Random_ID) == 1)
b <- hasins[is.na(hasins$U_MplusU_Ratio_Zen),]
b <- b[,1]
b$flagb <- rep(1,nrow(b))
# b contains list of IDs missing INS ratio
ab <- merge(a,b,by="Random_ID",all.x=TRUE, all.y=TRUE)
# people with both flaga and flagb are missing both measures and need to be excluded
ab$flaga[is.na(ab$flaga)] <- 0
ab$flagb[is.na(ab$flagb)] <- 0
ab$sum <- ab$flaga + ab$flagb
exclu <- ab[ab$sum==2,]
missboth <- length(unique(exclu$Random_ID))
alldata <- merge(alldata,exclu,by="Random_ID",all.x=TRUE,all.y=TRUE)
alldata <- alldata[is.na(alldata$sum),]

# who is missing T1D data?
missingt1d <- alldata[is.na(alldata$T1D),]
# no one is missing t1d status

# group status at each visit - but people can change groups
for(i in 1:nrow(alldata)) {
  alldata$group[i] <- ifelse(alldata$T1D[i]=="Yes","T1D",
                             ifelse(alldata$numpos[i]==0,"Negative",
                                    ifelse(alldata$numpos[i]==1,"Single ab positive","Multiple ab positive")))
}

#-------------------------------------------------------------------------------
# Tim's group status code.
# Separate table for diagnosis.
diagnosis.table <- alldata[,c("Random_ID","Draw_Date","group")]
# save a copy of this data to merge back with the final dx
allresults <- diagnosis.table
# Count diagnoses per subject
counts <- plyr::count(diagnosis.table,c("Random_ID","group"))
# Find participants with only one single ab+
single <- counts[counts$group == "Single ab positive",]
single <- single[single$freq < 2,"Random_ID"]
# Convert group to numeric
diagnosis.table$group <- 
  plyr::revalue(diagnosis.table$group,c("Negative" = 0,"Single ab positive" = 1,
                                        "Multiple ab positive" = 2,"T1D" = 3))
diagnosis.table$group <- as.numeric(diagnosis.table$group)
# Aggregate diagnosis table taking the maximum value.
diagnosis.table <- aggregate(group ~ Random_ID, data = diagnosis.table, max)
# Find the single ab+ who were incorrectly categorized, set back to 0.
recat <- single[
  which(single %in% diagnosis.table[diagnosis.table$group == 1,"Random_ID"])]
diagnosis.table[diagnosis.table$Random_ID %in% recat,"group"] <- 0
colnames(diagnosis.table) <- c("Random_ID","final.group")


# merge final group category with all results
allresults <- merge(allresults,diagnosis.table,by="Random_ID")
# make categorical variable for final group
allresults$final.group[allresults$final.group==0] <- "Negative"
allresults$final.group[allresults$final.group==1] <- "Single ab positive"
allresults$final.group[allresults$final.group==2] <- "Multiple ab positive"
allresults$final.group[allresults$final.group==3] <- "T1D"


# Merge the diagnosis back into the full table.
alldata <- merge(alldata,diagnosis.table,by = "Random_ID")
#-------------------------------------------------------------------------------
check <- alldata %>% distinct(Random_ID, .keep_all=TRUE)

# counting number of INS ratios and proinsulin ratios per person
#a <- nont1d %>% group_by(Random_ID) %>% filter(!is.na(Proinsulin_pM_L)) %>% summarize(count=n())
# range from 1 proinsulin ratio to 9
#b <- nont1d %>% group_by(Random_ID) %>% filter(!is.na(U_MplusU_Ratio_Zen)) %>% summarize(count=n())
# range from 1 INS ratio to 3

# find the first draw date
alldata <- alldata[order(alldata$Random_ID,alldata$Draw_Date),]
temp <- aggregate(alldata$Draw_Date,by=list(alldata$Random_ID),min)
names(temp) <- c("Random_ID","first_date")
temp$first_date <- as.Date(temp$first_date)
alldata <- merge(alldata,temp,by = "Random_ID")
alldata$Draw_Date <- as.Date(alldata$Draw_Date)
alldata$time <- alldata$Draw_Date - alldata$first_date
alldata$years <- alldata$time/365.25

# check range of years for non-missing pro-pep
a <- alldata[!is.na(alldata$Pro_pep),]
hist(as.numeric(a$years))
b <- alldata[!is.na(alldata$U_MplusU_Ratio_Zen),]
hist(as.numeric(b$years))

# spaghetti plots of outcomes
#qplot(years,Pro_pep,data=alldata,group=Random_ID,color=group,geom="line")
#qplot(years,U_MplusU_Ratio_Zen,data=alldata,group=Random_ID,color=group,geom="line")

# now need to determine which INS ratio/proinsulin ratio to use for cross-sectional analyses
# T1D: results closest to T1D onset 
alldata$Date_of_Diagnosis <- as.Date(mdy(alldata$Date_of_Diagnosis))
alldata$timec[alldata$final.group==3] <- abs(alldata$Draw_Date[alldata$final.group==3] - alldata$Date_of_Diagnosis[alldata$final.group==3])

# create labeled factor for final.group
#check <- alldata %>% distinct(Random_ID, .keep_all=TRUE)
#table(check$final.group)
alldata$final.group <- as.factor(alldata$final.group)
#check <- alldata %>% distinct(Random_ID, .keep_all=TRUE)
#table(check$final.group)
levels(alldata$final.group) = c("Negative","Single ab positive","Multiple ab positive","T1D")
#check <- alldata %>% distinct(Random_ID, .keep_all=TRUE)
#table(check$final.group)

# pull out T1D - if they have T1D at any time point, they are in the T1D group
t1d <- alldata[alldata$T1D == "Yes",]
t1d <- t1d[order(t1d$Random_ID, t1d$Draw_Date),]
# pull out nonT1D
nont1d <- alldata[alldata$T1D != "Yes",]
nont1d <- nont1d[order(nont1d$Random_ID, nont1d$Draw_Date),]
nneg <- length(unique(nont1d[nont1d$final.group=="Negative",]$Random_ID))
nsing <- length(unique(nont1d[nont1d$final.group=="Single ab positive",]$Random_ID))
nmult <- length(unique(nont1d[nont1d$final.group=="Multiple ab positive",]$Random_ID))
# total of 132 subjects, 15 in T1D and 117 nonT1D
# pull out records with non-missing Pro_pep
t1d_select_pro <- t1d[!is.na(t1d$Pro_pep),]
temp <-  aggregate(t1d_select_pro$timec,by=list(t1d_select_pro$Random_ID),min)
names(temp) <- c("Random_ID","time_from_diagnosis")
t1d_select_pro <- merge(t1d_select_pro,temp,by = "Random_ID")
t1d_select_pro <- t1d_select_pro[t1d_select_pro$time_from_diagnosis==t1d_select_pro$timec,]
# pull out records with nonmissing ins ratio
t1d_select_ins <- t1d[!is.na(t1d$U_MplusU_Ratio_Zen),]
length(unique(t1d_select_ins$Random_ID))
temp <-  aggregate(t1d_select_ins$timec,by=list(t1d_select_ins$Random_ID),min)
names(temp) <- c("Random_ID","time_from_diagnosis")
t1d_select_ins <- merge(t1d_select_ins,temp,by = "Random_ID")
t1d_select_ins <- t1d_select_ins[t1d_select_ins$time_from_diagnosis==t1d_select_ins$timec,]

# for negative, single ab positive, multiple ab positive: use most recent
# pull out records with nonmissing propep
nont1d_select_pro <- nont1d[!is.na(nont1d$Pro_pep),]
check <- nont1d_select_pro %>% distinct(Random_ID, .keep_all=TRUE)
# select the most recent record for each person
temp <- aggregate(nont1d_select_pro$Draw_Date,by=list(nont1d_select_pro$Random_ID),max)
names(temp) <- c("Random_ID","most_recent_draw")
temp$most_recent_draw <- as.Date(temp$most_recent_draw)
nont1d_select_pro <- merge(nont1d_select_pro,temp,by = "Random_ID")
nont1d_select_pro <- nont1d_select_pro[nont1d_select_pro$Draw_Date==nont1d_select_pro$most_recent_draw,]
# pull out records with nonmissing ins ratio
nont1d_select_ins <-nont1d[!is.na(nont1d$U_MplusU_Ratio_Zen),]
# select the most recent record for each person
temp <- aggregate(nont1d_select_ins$Draw_Date,by=list(nont1d_select_ins$Random_ID),max)
names(temp) <- c("Random_ID","most_recent_draw")
temp$most_recent_draw <- as.Date(temp$most_recent_draw)
nont1d_select_ins <- merge(nont1d_select_ins,temp,by = "Random_ID")
nont1d_select_ins <- nont1d_select_ins[nont1d_select_ins$Draw_Date==nont1d_select_ins$most_recent_draw,]

# combine the results for the pro-pep analysis
t1d_select_pro$most_recent_draw <- NA
nont1d_select_pro$time_from_diagnosis <- NA
allpro <- rbind(nont1d_select_pro,t1d_select_pro)
# combine the results for the INS analysis
t1d_select_ins$most_recent_draw <- NA
nont1d_select_ins$time_from_diagnosis <- NA
allins <- rbind(nont1d_select_ins,t1d_select_ins)

# for table 1, need the draw date corresponding to the proins and INS ratios used in cross sectional analyses
datepro <- allpro[c("Random_ID","Draw_Date")]
names(datepro) <- c("Random_ID","Date_proins")
dateINS <- allins[c("Random_ID","Draw_Date")]
names(dateINS) <- c("Random_ID","Date_INS")
getdates <- merge(datepro,dateINS,by="Random_ID",all.x=TRUE, all.y=TRUE)
# keep the latest date to find A1c and age
getdates$tab1_date <- pmax(getdates$Date_INS,getdates$Date_proins,na.rm=TRUE)
getdates$tab1flag <- rep(1,nrow(getdates))
getdates$Draw_Date <- getdates$tab1_date
tab1data <- merge(alldata,getdates,by=c("Random_ID","Draw_Date"))
tab1data$age <- floor(tab1data$DOB %--% tab1data$Draw_Date / dyears(1))

# compare unmethylated INS ratios between T1D, Ab-, single Ab+, multiple Ab+
allins$l_U_MplusU_Ratio_Zen=log(allins$U_MplusU_Ratio_Zen)
# lm.ins = with(allins, lm(U_MplusU_Ratio_Zen ~ as.factor(final.group))) 
# summary(lm.ins)
# anova(lm.ins)
# plot(lm.ins)
# lm.ins <- aov(lm.ins)
# summary(lm.ins)
# anova(lm.ins)
# TukeyHSD(lm.ins)
# log transformed
lm.ins = with(allins, lm(l_U_MplusU_Ratio_Zen ~ as.factor(final.group))) 
summary(lm.ins)
anova(lm.ins)
lm.ins <- aov(lm.ins)
summary(lm.ins)
anova(lm.ins)
TukeyHSD(lm.ins)

tab_ins <- final_table(allins,'U_MplusU_Ratio_Zen',group=allins$final.group,margin=2,single=0,ron=2)

# do ins ratios correlate with biochemical and ECL antibody levels
# ++++++++++++++++++++++++++++
# flattenCorrMatrix
# ++++++++++++++++++++++++++++
# cormat : matrix of the correlation coefficients
# pmat : matrix of the correlation p-values
flattenCorrMatrix <- function(cormat, pmat) {
  ut <- upper.tri(cormat)
  data.frame(
    row = rownames(cormat)[row(cormat)[ut]],
    column = rownames(cormat)[col(cormat)[ut]],
    cor  =(cormat)[ut],
    p = pmat[ut]
    )
}
corr_ins <- rcorr(as.matrix(allins[, c( "U_MplusU_Ratio_Zen","gadz","ia2z","mIAA","ZnT8","ICA",
                                        "ECL_GADA","ECL_IA2A","ECL_IAA")])
                  ,type = "spearman")
corr_ins <- flattenCorrMatrix(corr_ins$r, corr_ins$P)
corr_ins <- corr_ins[corr_ins$row=="U_MplusU_Ratio_Zen",]

# compare proinsulin ratios between T1D, Ab-, single Ab+, multiple Ab+
allpro$l_Pro_pep=log(allpro$Pro_pep)
# lm.pro = with(allpro, lm(Pro_pep ~ as.factor(final.group))) 
# summary(lm.pro)
# anova(lm.pro)
# plot(lm.pro)
# lm.pro <- aov(lm.pro)
# summary(lm.pro)
# anova(lm.pro)
# TukeyHSD(lm.pro)
# log transformed
lm.pro = with(allpro, lm(l_Pro_pep ~ as.factor(final.group))) 
summary(lm.pro)
anova(lm.pro)
lm.pro <- aov(lm.pro)
summary(lm.pro)
anova(lm.pro)
TukeyHSD(lm.pro)
tab_pro <- final_table(allpro,'Pro_pep',group=as.factor(allpro$final.group),margin=2,single=0,ron=2)

# do proinsulin ratios correlate with biochemical and ECL antibody levels
corr_pro <- rcorr(as.matrix(allpro[, c( "Pro_pep","gadz","ia2z","mIAA","ZnT8","ICA",
                                        "ECL_GADA","ECL_IA2A","ECL_IAA")])
                  ,type = "spearman")
corr_pro <- flattenCorrMatrix(corr_pro$r, corr_pro$P)
corr_pro <- corr_pro[corr_pro$row=="Pro_pep",]

# table 1 - include age at last visit, gender, race/eth, DPT1 score, HLA, A1c
label(tab1data$age)="Age (yrs)"
label(tab1data$HLA)="HLA"
label(tab1data$HbA1c)="HbA1c (%)"
label(tab1data$race_eth)="Race/ethnicity"
label(tab1data$DR3DR4)="DR3 DR4"
tab1data$DR3DR4 <- as.factor(tab1data$DR3DR4)
t1 <- final_table(tab1data,c("age","Gender","race_eth","DPT1Risk","DR3","DR4","DR3DR4","HLA","HbA1c"),group=tab1data$final.group,margin=2,single=0,ron=1)


```

## Background

The purpose of this analysis is to test whether unmethylated INS ratios and proinsulin ratios differ between 4 groups of TrialNet subjects: T1D, multiple antibody positive, single antibody positive, and negative.  The correlation of unmethylated INS ratios and proinsulin ratios with biochemical and ECL antibody levels was also examined.

## Methods

There were `r obslong` records in the initial longitudinal dataset, and `r nppt` participants in the participant database.

The following data manipulation and cleaning steps were performed:

1) Fasting C-peptide in pM/L was calculated as fasting C-peptide*331.09.
2) Proinsulin ratios were calculated as proinsulin (pM/L) / C-peptide (pM/L) *100
3) There were `r nrow(misstwo)` records deleted that were missing both mIAA and ICA512/IA2H results.
4) There were an additional `r nrow(missmiaa)` records deleted that were missing just mIAA.
5) Antibody results were classifed as positive/negative using the following cutoff values:
+ mIAA, 0.01
+ GAD65, 0.032
+ GAD65H, 20
+ ICA512, 0.049
+ IA_2H, 5
+ ZnT8, 0.02
6) The categorical variables (positive/negative) for GAD65 and GAD65H were combined into one variable, as were the categorical variables for ICA512 and IA_2H.
7) The number of positive antibodies at each visit was calculated as the sum of ICA512/IA2H, GAD65/GAD65H, mIAA, and ZnT8.
8) For analyses using the continuous value of the antibodies, GAD65 and GAD65H were combined by calculating the Z-score of each variable and then combining the Z-scores into one variable.  The same procedure was followed for ICA512/IA2H.
9) There were `r missboth` subjects who did not have any proinsulin ratio results or any unmethylated INS ratios who were excluded from analysis.   
10) Participants were classified in the T1D group if they ever had a diagnosis of T1D.  
11) If a participant was not in the T1D group, they were classfied by antibody status as below.  This resulted in `r length(unique(t1d$Random_ID))` participants in the T1D group, `r nmult` participants in the multiple antibody positive group, `r nsing` participants in the single antibody positive group, and `r nneg` participants in the negative group.
+ If they were multiple antibody positive at any visit, they were classified as multiple antibody positive.
+ If they were never multiple antibody positive, but were single antibody positive at 2 or more visits, they were classified as single antibody positive.
+ Otherwise, they were classified as antibody negative.
12) For cross-sectional analyses, unmethylated INS ratios and proinsulin ratios were selected as follows:
+ For the T1D group, the result closest to the date of diagnosis was used.
+ For all other groups, the most recent result was used.

## Results

Table 1 presents descriptive statistics for demographic and clinical variables by group.

```{r, results='asis',tidy=TRUE, echo=FALSE}
kable(t1,caption="Table 1.  Descriptive statistics.  Asterisks indicate that Fisher's exact test was used instead of the chi-square test due to small expected cell value.")
```
\

Tables 2 and 3 show the results of the group comparisons for unmethylated INS ratio and proinsulin ratio.

```{r, results='asis',tidy=TRUE, echo=FALSE}
kable(tab_ins,caption="Table 2.  Unmethylated insulin ratios.")
```
\
```{r, results='asis',tidy=TRUE, echo=FALSE}
kable(tab_pro,caption="Table 3.  Proinsulin ratios.")
```
\

Tables 4 and 5 show the correlations of unmethylated INS and proinsulin ratios with continuous antibody levels at the timepoint selected for cross-sectional analysis.
```{r, results='asis',tidy=TRUE, echo=FALSE}
kable(corr_ins,caption="Table 4.  Correlation of unmethylated insulin ratios with antibody levels.")
```
\
```{r, results='asis',tidy=TRUE, echo=FALSE}
kable(corr_pro,caption="Table 5.  Correlation of proinsulin ratios with antibody levels.")
```
\
