---
title: "TrialNet INS ratio analyses"
author: "Laura Pyle"
date: "November 11, 2018"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#library(sqldf)
library(dplyr)
library(lubridate)
library(graphics)
library(ggplot2)
library(Hmisc)
library(Table1)
library(knitr)

# read in the data
longitudinal_visit_info <- read.csv("\\\\ucdenver.pvt\\som-home\\pylell\\BDC\\Steck\\Trialnet progressors analysis\\Fourth analysis - demethylated insulin with updated dataset\\Data\\longitudinal 091818.csv")
ParticipantInfo_risk <- read.csv("\\\\ucdenver.pvt\\som-home\\pylell\\BDC\\Steck\\Trialnet progressors analysis\\Fourth analysis - demethylated insulin with updated dataset\\Data\\Participant Info risk 091818.csv")

# get rid of duplicates
ParticipantInfo_risk <- unique(ParticipantInfo_risk)
# no duplicates

obslong <- nrow(longitudinal_visit_info)
nppt <- nrow(ParticipantInfo_risk)

# merge participant info and longitudinal data
keep_ppt <- data.frame(ParticipantInfo_risk$TrialNet,
                       ParticipantInfo_risk$T1D,
                       ParticipantInfo_risk$Date.of.Diagnosis) 
colnames(keep_ppt) <- c("Random_ID","T1D","Date_of_Diagnosis")
keep_long <- data.frame(longitudinal_visit_info$PID,
                        longitudinal_visit_info$Draw.Date,
                        longitudinal_visit_info$GAD65,
                        longitudinal_visit_info$GAD65H,
                        longitudinal_visit_info$ICA512,
                        longitudinal_visit_info$IA2H,
                        longitudinal_visit_info$mIAA,
                        longitudinal_visit_info$ZnT8,
                        longitudinal_visit_info$ICA,
                        longitudinal_visit_info$U.per.MplusU.Ratio.Zen,
                        longitudinal_visit_info$ECL_GADA,
                        longitudinal_visit_info$ECL_IA.2A,
                        longitudinal_visit_info$ECL_IAA,
                        longitudinal_visit_info$Proinsulin_pM.L,
                        longitudinal_visit_info$PEP0)
colnames(keep_long) <- c("Random_ID","Draw_Date","GAD65","GAD65H","ICA512","IA_2H","mIAA","ZnT8","ICA","U_MplusU_Ratio_Zen","ECL_GADA","ECL_IA2A","ECL_IAA","Proinsulin_pM_L","PEP0")

# convert draw date to date
#keep_long$Draw_Date <- mdy_hms(as.character(keep_long$Draw_Date))
#keep_long$Draw_Date <- as.POSIXlt(as.character(keep_long$Draw_Date),format="%m%d%Y %H%M")
keep_long$Draw_Date <- mdy_hm(as.character(keep_long$Draw_Date))

# calculate fasting Cpeptide in pM/L
keep_long$PEP0_pM_L <- keep_long$PEP0*331.09

# calculate proinsulin/Cpep ratio
keep_long$Pro_pep <- keep_long$Proinsulin_pM_L/keep_long$PEP0_pM_L*100

# merge  datasets
alldata <- merge(keep_ppt,keep_long,by="Random_ID",all.x=F,all.y=F)
N <- length(unique(alldata$Random_ID))

# labels
label(alldata$U_MplusU_Ratio_Zen)='INS ratio'
label(alldata$Pro_pep)='Proinsulin ratio'

# first code each antibody as positive negative
for(i in 1:nrow(alldata)){
  ###coding for miaa
  a <-ifelse(is.na(alldata$mIAA[i]), 
             NA,
             ifelse(alldata$mIAA[i] <= 0.01, 0, 1)
  )
  alldata$miaa_cat[i]<- a
  ###coding for first form of gad65
  b <-ifelse(is.na(alldata$GAD65[i]), 
             NA,
             ifelse(alldata$GAD65[i] <= 0.032, 0, 1)
  )
  alldata$gad65_cat[i]<- b
  ###coding for second form of gad65
  c<-ifelse(is.na(alldata$GAD65H[i]), 
            NA,
            ifelse(alldata$GAD65H[i] <= 20, 0, 1)
  )
  alldata$gad65h_cat[i]<- c
  
  ###coding for first form of ica512/ia2
  d<-ifelse(is.na(alldata$ICA512[i]), 
            NA,
            ifelse(alldata$ICA512[i] <= 0.049, 0, 1)
  )
  alldata$ica512_cat[i]<- d
  ###coding for second form of ica512/ia2
  e<-ifelse(is.na(alldata$IA_2H[i]), 
            NA,
            ifelse(alldata$IA_2H[i] <= 5, 0, 1)
  )
  alldata$ia2h_cat[i]<- e
  ###coding for znt8
  f<-ifelse(is.na(alldata$ZnT8[i]), 
            NA,
            ifelse(alldata$ZnT8[i] <= 0.02, 0, 1)
  )
  alldata$znt8_cat[i]<- f
}

#####combining both forms of gad65 tests
for(i in 1:nrow(alldata)){
  test <- sum(alldata$ica512_cat[i],alldata$ia2h_cat[i],na.rm=TRUE)
  de <- ifelse(is.na(alldata$ica512_cat[i]) & is.na(alldata$ia2h_cat[i]),
               NA,
               ifelse(test>=1,1,0))
  alldata$ica512_ia2_cat[i]<- de    
  test2 <- sum(alldata$gad65_cat[i],alldata$gad65h_cat[i],na.rm=TRUE)
  bc <- ifelse(is.na(alldata$gad65_cat[i]) & is.na(alldata$gad65h_cat[i]),
               1,
               ifelse(test2>=1,1,0))
  alldata$GAD65_combo_cat[i]<- bc   
}  

# if all ab results missing, delete the record
misstwo <- alldata[is.na(alldata$mIAA) & is.na(alldata$ica512_ia2_cat),]
alldata <- alldata[!is.na(alldata$mIAA) | !is.na(alldata$miaa_cat) | !is.na(alldata$ica512_ia2_cat),]
# 36 records deleted

# check if anyone is missing GAD, ICA/IA2, or mIAA
missing <- alldata[is.na(alldata$GAD65_combo_cat) | is.na(alldata$miaa_cat) | is.na(alldata$ica512_ia2_cat),]
write.csv(missing,file="missing.csv") 

# get rid of visits with missing mIAA data
missmiaa <- alldata[is.na(alldata$GAD65_combo_cat) | is.na(alldata$miaa_cat) | is.na(alldata$ica512_ia2_cat),]
alldata <- alldata[!is.na(alldata$GAD65_combo_cat) & !is.na(alldata$miaa_cat) & !is.na(alldata$ica512_ia2_cat),]
# 40 records deleted

# calculate number of positive ab at each visit
for(i in 1:nrow(alldata)){
  alldata$numpos[i] <- sum(alldata$ica512_ia2_cat[i],alldata$GAD65_combo_cat[i],alldata$miaa_cat[i],alldata$znt8_cat[i],na.rm=T)
}

# calculate z-scores for GAD and IA2
alldata$gad65z <- (alldata$GAD65 - mean(alldata$GAD65,na.rm = TRUE))/sd(alldata$GAD65,na.rm = TRUE)
alldata$gad65hz <- (alldata$GAD65H - mean(alldata$GAD65H,na.rm = TRUE))/sd(alldata$GAD65H,na.rm = TRUE)
alldata$gadz <- ifelse(!is.na(alldata$GAD65),alldata$gad65z,alldata$gad65hz)
alldata$ica512z <- (alldata$ICA512 - mean(alldata$ICA512,na.rm = TRUE))/sd(alldata$ICA512,na.rm = TRUE)
alldata$ia2hz <- (alldata$IA_2H - mean(alldata$IA_2H,na.rm = TRUE))/sd(alldata$IA_2H,na.rm = TRUE)
alldata$ia2z <- ifelse(!is.na(alldata$ICA512),alldata$ica512z,alldata$ia2hz)

# need to find out if anyone is missing proinsulin ratio at all visits
sortds <- alldata[order(alldata$Random_ID,desc(alldata$Proinsulin_pM_L)),]
hasproins <- sortds %>% group_by(Random_ID) %>% filter(row_number(Random_ID) == 1)
a <- hasproins[is.na(hasproins$Proinsulin_pM_L),]
a <- a[,1]
misspro <- nrow(a)
a$flag <- rep(1,nrow(a))
# 12 subjects missing proinsulin - need to merge this back with original dataset and take those people out
# keep only those IDs that are in both dataframes
alldata <- merge(alldata,a,all.x=TRUE,by="Random_ID")
alldata <- alldata[is.na(alldata$flag),]

# who is missing T1D data?
missingt1d <- alldata[is.na(alldata$T1D),]
# no one is missing t1d status

# group status at each visit - but people can change groups
for(i in 1:nrow(alldata)) {
  alldata$group[i] <- ifelse(alldata$T1D[i]=="Yes","T1D",
                             ifelse(alldata$numpos[i]==0,"Negative",
                                    ifelse(alldata$numpos[i]==1,"Single ab positive","Multiple ab positive")))
}

#-------------------------------------------------------------------------------
# Tim's group status code.
# Separate table for diagnosis.
diagnosis.table <- alldata[,c("Random_ID","Draw_Date","group")]
# save a copy of this data to merge back with the final dx
allresults <- diagnosis.table
# Count diagnoses per subject
counts <- plyr::count(diagnosis.table,c("Random_ID","group"))
# Find participants with only one single ab+
single <- counts[counts$group == "Single ab positive",]
single <- single[single$freq < 2,"Random_ID"]
# Convert group to numeric
diagnosis.table$group <- 
  plyr::revalue(diagnosis.table$group,c("Negative" = 0,"Single ab positive" = 1,
                                        "Multiple ab positive" = 2,"T1D" = 3))
diagnosis.table$group <- as.numeric(diagnosis.table$group)
# Aggregate diagnosis table taking the maximum value.
diagnosis.table <- aggregate(group ~ Random_ID, data = diagnosis.table, max)
# Find the single ab+ who were incorrectly categorized, set back to 0.
recat <- single[
  which(single %in% diagnosis.table[diagnosis.table$group == 1,"Random_ID"])]
diagnosis.table[diagnosis.table$Random_ID %in% recat,"group"] <- 0
colnames(diagnosis.table) <- c("Random_ID","final.group")


# merge final group category with all results
allresults <- merge(allresults,diagnosis.table,by="Random_ID")
# make categorical variable for final group
allresults$final.group[allresults$final.group==0] <- "Negative"
allresults$final.group[allresults$final.group==1] <- "Single ab positive"
allresults$final.group[allresults$final.group==2] <- "Multiple ab positive"
allresults$final.group[allresults$final.group==3] <- "T1D"


# Merge the diagnosis back into the full table.
alldata <- merge(alldata,diagnosis.table,by = "Random_ID")
#-------------------------------------------------------------------------------

# counting number of INS ratios and proinsulin ratios per person
#a <- nont1d %>% group_by(Random_ID) %>% filter(!is.na(Proinsulin_pM_L)) %>% summarize(count=n())
# range from 1 proinsulin ratio to 9
#b <- nont1d %>% group_by(Random_ID) %>% filter(!is.na(U_MplusU_Ratio_Zen)) %>% summarize(count=n())
# range from 1 INS ratio to 3

# find the first draw date
alldata <- alldata[order(alldata$Random_ID,alldata$Draw_Date),]
temp <- aggregate(alldata$Draw_Date,by=list(alldata$Random_ID),min)
names(temp) <- c("Random_ID","first_date")
temp$first_date <- as.Date(temp$first_date)
alldata <- merge(alldata,temp,by = "Random_ID")
alldata$Draw_Date <- as.Date(alldata$Draw_Date)
alldata$time <- alldata$Draw_Date - alldata$first_date
alldata$years <- alldata$time/365.25

# check range of years for non-missing pro-pep
a <- alldata[!is.na(alldata$Pro_pep),]
hist(as.numeric(a$years))
b <- alldata[!is.na(alldata$U_MplusU_Ratio_Zen),]
hist(as.numeric(b$years))

# spaghetti plots of outcomes
qplot(years,Pro_pep,data=alldata,group=Random_ID,color=group,geom="line")
qplot(years,U_MplusU_Ratio_Zen,data=alldata,group=Random_ID,color=group,geom="line")

# now need to determine which INS ratio/proinsulin ratio to use for cross-sectional analyses
# T1D: results closest to T1D onset 
alldata$Date_of_Diagnosis <- as.Date(mdy(alldata$Date_of_Diagnosis))
alldata$timec[alldata$final.group==3] <- abs(alldata$Draw_Date[alldata$final.group==3] - alldata$Date_of_Diagnosis[alldata$final.group==3])

# create labeled factor for final.group
#check <- alldata %>% distinct(Random_ID, .keep_all=TRUE)
#table(check$final.group)
alldata$final.group <- as.factor(alldata$final.group)
#check <- alldata %>% distinct(Random_ID, .keep_all=TRUE)
#table(check$final.group)
levels(alldata$final.group) = c("Negative","Single ab positive","Multiple ab positive","T1D")
#check <- alldata %>% distinct(Random_ID, .keep_all=TRUE)
#table(check$final.group)

# pull out T1D - if they have T1D at any time point, they are in the T1D group
t1d <- alldata[alldata$T1D == "Yes",]
t1d <- t1d[order(t1d$Random_ID, t1d$Draw_Date),]
# pull out nonT1D
nont1d <- alldata[alldata$T1D != "Yes",]
nont1d <- nont1d[order(nont1d$Random_ID, nont1d$Draw_Date),]
nneg <- length(unique(nont1d[nont1d$final.group==0,]$Random_ID))
nsing <- length(unique(nont1d[nont1d$final.group==1,]$Random_ID))
nmult <- length(unique(nont1d[nont1d$final.group==2,]$Random_ID))
# total of 132 subjects, 15 in T1D and 117 nonT1D
# pull out records with non-missing Pro_pep
t1d_select_pro <- t1d[!is.na(t1d$Pro_pep),]
temp <-  aggregate(t1d_select_pro$timec,by=list(t1d_select_pro$Random_ID),min)
names(temp) <- c("Random_ID","time_from_diagnosis")
t1d_select_pro <- merge(t1d_select_pro,temp,by = "Random_ID")
t1d_select_pro <- t1d_select_pro[t1d_select_pro$time_from_diagnosis==t1d_select_pro$timec,]
# pull out records with nonmissing ins ratio
t1d_select_ins <- t1d[!is.na(t1d$U_MplusU_Ratio_Zen),]
length(unique(t1d_select_ins$Random_ID))
temp <-  aggregate(t1d_select_ins$timec,by=list(t1d_select_ins$Random_ID),min)
names(temp) <- c("Random_ID","time_from_diagnosis")
t1d_select_ins <- merge(t1d_select_ins,temp,by = "Random_ID")
t1d_select_ins <- t1d_select_ins[t1d_select_ins$time_from_diagnosis==t1d_select_ins$timec,]

# for negative, single ab positive, multiple ab positive: use most recent
# pull out records with nonmissing propep
nont1d_select_pro <- nont1d[!is.na(nont1d$Pro_pep),]
# select the most recent record for each person
temp <- aggregate(nont1d_select_pro$Draw_Date,by=list(nont1d_select_pro$Random_ID),max)
names(temp) <- c("Random_ID","most_recent_draw")
temp$most_recent_draw <- as.Date(temp$most_recent_draw)
nont1d_select_pro <- merge(nont1d_select_pro,temp,by = "Random_ID")
nont1d_select_pro <- nont1d_select_pro[nont1d_select_pro$Draw_Date==nont1d_select_pro$most_recent_draw,]
# pull out records with nonmissing ins ratio
nont1d_select_ins <-nont1d[!is.na(nont1d$U_MplusU_Ratio_Zen),]
# select the most recent record for each person
temp <- aggregate(nont1d_select_ins$Draw_Date,by=list(nont1d_select_ins$Random_ID),max)
names(temp) <- c("Random_ID","most_recent_draw")
temp$most_recent_draw <- as.Date(temp$most_recent_draw)
nont1d_select_ins <- merge(nont1d_select_ins,temp,by = "Random_ID")
nont1d_select_ins <- nont1d_select_ins[nont1d_select_ins$Draw_Date==nont1d_select_ins$most_recent_draw,]


# # negative: use most recent
# neg <- alldata[alldata$final.group==0,]
# neg <- neg[order(neg$Random_ID, neg$Draw_Date),]
# # pull out records with nonmissing propep
# neg_select_pro <- neg[!is.na(neg$Pro_pep),]
# # select the most recent record for each person
# temp <- aggregate(neg_select_pro$Draw_Date,by=list(neg_select_pro$Random_ID),max)
# names(temp) <- c("Random_ID","most_recent_draw")
# temp$most_recent_draw <- as.Date(temp$most_recent_draw)
# neg_select_pro <- merge(neg_select_pro,temp,by = "Random_ID")
# length(unique(neg_select_pro$Random_ID))
# neg_select_pro <- neg_select_pro[neg_select_pro$Draw_Date==neg_select_pro$most_recent_draw,]
# # pull out records with nonmissing ins ratio
# neg_select_ins <-neg[!is.na(neg$U_MplusU_Ratio_Zen),]
# # select the most recent record for each person
# temp <- aggregate(neg_select_ins$Draw_Date,by=list(neg_select_ins$Random_ID),max)
# names(temp) <- c("Random_ID","most_recent_draw")
# temp$most_recent_draw <- as.Date(temp$most_recent_draw)
# neg_select_ins <- merge(neg_select_ins,temp,by = "Random_ID")
# neg_select_ins <- neg_select_ins[neg_select_ins$Draw_Date==neg_select_ins$most_recent_draw,]
# 
# # Multiple ab positive: results closest to multiple ab positivity
# # if anyone has more than one set of results that meets the above criteria, take the most recent result
# # first need to find the first date of multiple ab positivity
# mult <- alldata[alldata$final.group==2,]
# mult <- mult[order(mult$Random_ID, mult$Draw_Date),]
# mult_find_first <- mult[mult$group=="Multiple ab positive",]
# mult_find_first <- aggregate(mult_find_first$Draw_Date,by=list(mult_find_first$Random_ID),min)
# names(mult_find_first) <- c("Random_ID","first_mult_date")
# mult <- merge(mult,mult_find_first,by="Random_ID")
# # calculate difference between draw date and date of multiple ab positivity
# mult$timec <- mult$Draw_Date - mult$first_mult_date
# View(mult[c("Random_ID","group","final.group","Draw_Date","first_mult_date","timec")])
# # pull out records with non-missing Pro_pep
# mult_select_pro <- mult[!is.na(mult$Pro_pep),]
# length(unique(mult_select_pro$Random_ID))
# mult_select_pro <- mult_select_pro[abs(mult_select_pro$timec)<=60,]
# length(unique(mult_select_pro$Random_ID))
# 
# # single ab positive: results at time of confirmation
# # if anyone has more than one set of results that meets the above criteria, take the most recent result

# combine the results for the pro-pep analysis
t1d_select_pro$most_recent_draw <- NA
nont1d_select_pro$time_from_diagnosis <- NA
allpro <- rbind(nont1d_select_pro,t1d_select_pro)
# combine the results for the INS analysis
t1d_select_ins$most_recent_draw <- NA
nont1d_select_ins$time_from_diagnosis <- NA
allins <- rbind(nont1d_select_ins,t1d_select_ins)

# compare unmethylated INS ratios between T1D, Ab-, single Ab+, multiple Ab+
allins$l_U_MplusU_Ratio_Zen=log(allins$U_MplusU_Ratio_Zen)
# lm.ins = with(allins, lm(U_MplusU_Ratio_Zen ~ as.factor(final.group))) 
# summary(lm.ins)
# anova(lm.ins)
# plot(lm.ins)
# lm.ins <- aov(lm.ins)
# summary(lm.ins)
# anova(lm.ins)
# TukeyHSD(lm.ins)
# log transformed
lm.ins = with(allins, lm(l_U_MplusU_Ratio_Zen ~ as.factor(final.group))) 
summary(lm.ins)
anova(lm.ins)
lm.ins <- aov(lm.ins)
summary(lm.ins)
anova(lm.ins)
TukeyHSD(lm.ins)

tab_ins <- final_table(allins,'U_MplusU_Ratio_Zen',group=allins$final.group,margin=2,single=0,ron=2)

# do ins ratios correlate with biochemical and ECL antibody levels
# ++++++++++++++++++++++++++++
# flattenCorrMatrix
# ++++++++++++++++++++++++++++
# cormat : matrix of the correlation coefficients
# pmat : matrix of the correlation p-values
flattenCorrMatrix <- function(cormat, pmat) {
  ut <- upper.tri(cormat)
  data.frame(
    row = rownames(cormat)[row(cormat)[ut]],
    column = rownames(cormat)[col(cormat)[ut]],
    cor  =(cormat)[ut],
    p = pmat[ut]
    )
}
corr_ins <- rcorr(as.matrix(allins[, c( "U_MplusU_Ratio_Zen","gadz","ia2z","mIAA","ZnT8","ICA",
                                        "ECL_GADA","ECL_IA2A","ECL_IAA")])
                  ,type = "spearman")
corr_ins <- flattenCorrMatrix(corr_ins$r, corr_ins$P)
corr_ins <- corr_ins[corr_ins$row=="U_MplusU_Ratio_Zen",]

# compare proinsulin ratios between T1D, Ab-, single Ab+, multiple Ab+
allpro$l_Pro_pep=log(allpro$Pro_pep)
# lm.pro = with(allpro, lm(Pro_pep ~ as.factor(final.group))) 
# summary(lm.pro)
# anova(lm.pro)
# plot(lm.pro)
# lm.pro <- aov(lm.pro)
# summary(lm.pro)
# anova(lm.pro)
# TukeyHSD(lm.pro)
# log transformed
lm.pro = with(allpro, lm(l_Pro_pep ~ as.factor(final.group))) 
summary(lm.pro)
anova(lm.pro)
lm.pro <- aov(lm.pro)
summary(lm.pro)
anova(lm.pro)
TukeyHSD(lm.pro)
tab_pro <- final_table(allpro,'Pro_pep',group=as.factor(allpro$final.group),margin=2,single=0,ron=2)

# do proinsulin ratios correlate with biochemical and ECL antibody levels
corr_pro <- rcorr(as.matrix(allpro[, c( "Pro_pep","gadz","ia2z","mIAA","ZnT8","ICA",
                                        "ECL_GADA","ECL_IA2A","ECL_IAA")])
                  ,type = "spearman")
corr_pro <- flattenCorrMatrix(corr_pro$r, corr_pro$P)
corr_pro <- corr_pro[corr_pro$row=="Pro_pep",]


```

## Background

The purpose of this analysis is to test whether unmethylated INS ratios and proinsulin ratios differ between 4 groups of TrialNet subjects: T1D, multiple antibody positive, single antibody positive, and negative.  The correlation of unmethylated INS ratios and proinsulin ratios with biochemical and ECL antibody levels was also examined.

## Methods

There were `r obslong` records in the initial longitudinal dataset, and `r nppt` participants in the participant database.

The following data manipulation and cleaning steps were performed:

1) Fasting C-peptide in pM/L was calculated as fasting C-peptide*331.09.
2) Proinsulin ratios were calculated as proinsulin (pM/L) / C-peptide (pM/L) *100
3) There were `r nrow(misstwo)` records deleted that were missing both mIAA and ICA512/IA2H results.
4) There were an additional `r nrow(missmiaa)` records deleted that were missing just mIAA.
5) Antibody results were classifed as positive/negative using the following cutoff values:
+ mIAA, 0.01
+ GAD65, 0.032
+ GAD65H, 20
+ ICA512, 0.049
+ IA_2H, 5
+ ZnT8, 0.02
6) The categorical variables (positive/negative) for GAD65 and GAD65H were combined into one variable, as were the categorical variables for ICA512 and IA_2H.
7) The number of positive antibodies at each visit was calculated as the sum of ICA512/IA2H, GAD65/GAD65H, mIAA, and ZnT8.
8) For analyses using the continuous value of the antibodies, GAD65 and GAD65H were combined by calculating the Z-score of each variable and then combining the Z-scores into one variable.  The same procedure was followed for ICA512/IA2H.
9) There were `r misspro` subjects who did not have any proinsulin results who were excluded from analysis.
10) Participants were classified in the T1D group if they ever had a diagnosis of T1D.  
11) If a participant was not in the T1D group, they were classfied by antibody status as below.  This resulted in `r length(unique(t1d$Random_ID))` participants in the T1D group, `r nmult` participants in the multiple antibody positive group, `r nsing` participants in the single antibody positive group, and `r nneg` participants in the negative group.
+ If they were multiple antibody positive at any visit, they were classified as multiple antibody positive.
+ If they were never multiple antibody positive, but were single antibody positive at 2 or more visits, they were classified as single antibody positive.
+ Otherwise, they were classified as antibody negative.
12) For cross-sectional analyses, unmethylated INS ratios and proinsulin ratios were selected as follows:
+ For the T1D group, the result closest to the date of diagnosis was used.
+ For all other groups, the most recent result was used.

## Results

Tables 2 and 3 show the results of the group comparisons for unmethylated INS ratio and proinsulin ratio.

```{r, results='asis',tidy=TRUE, echo=FALSE}
kable(tab_ins,caption="Table 2.  Unmethylated insulin ratios.")
```
\
```{r, results='asis',tidy=TRUE, echo=FALSE}
kable(tab_pro,caption="Table 3.  Proinsulin ratios.")
```
\

Tables 4 and 5 show the correlations of unmethylated INS and proinsulin ratios with continuous antibody levels.
```{r, results='asis',tidy=TRUE, echo=FALSE}
kable(corr_ins,caption="Table 4.  Correlation of unmethylated insulin ratios with antibody levels.")
```
\
```{r, results='asis',tidy=TRUE, echo=FALSE}
kable(corr_pro,caption="Table 5.  Correlation of proinsulin ratios with antibody levels.")
```
\
